---
title: "Workshops"
output: html_document
date: "2023-05-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages("tidyverse")
library(tidyverse)
```

The mpg data frame found in ggplot2 is a dataset with observations (234 rows of data) and variables (in the columns)
```{r}
mpg
#disp = engine size, hwy = fuel efficiency in miles per gallon 
```
## Plotting with ggplot2
#ggplot template 
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```
## Mapping aesthetics 
Change point colour by class
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```
Change pint size by class
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```
Using size for a discrete value is not advised. Don’t plot a variable which is not continuous to a continuous aesthetic like size

Change point transparency by class (alpha)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```
Change point shape by class.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = displ < 5)
```
## Facet and Panel Plots 
To facet your plot using a single variable, use facet_wrap(). Only use facet_wrap() for discrete variables.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```
If you want to plot more than one variable, then use facet_grid(). Use ~ to split them up.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
#categorzing the plots into drive and cylindical size 
```
Excersize 1: Read ?facet_wrap
```{r}
?facet_wrap
```
* nrow & ncol determine the number of rows and columns 
* as.table is set to default TRUE, meaning that the facets are laid out like a table with highest values at the bottom-right
* dir determines the direction. EIther "h" for horizontal, or the default "v" for vertical. 

## Fitting Simple Lines
Using visual objects to represent the data. 

To display data as points, use geom_point()
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```
To display the same data as a smooth line fit through the points use geom_smooth()
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
You can change the line type and can use a variable to control it.
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```
Here we will use the group argument to show that our data is in fact grouped - it won’t add a new legend but can easily be used to vary your plots by the discrete variable.

Set the group aesthetic to a categorical variable to draw multiple objects
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```
Change the color of each line based on drv value
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  )
```
Here we will plot multiple geoms on the single plot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
Global mappings are applied to every subsequent geom 
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```
Use mappings in specific layers to enable the display of different aesthetics in different layers. Note how the line is not styled by class but you can style the points by themselves.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```
The same goes if you want to specify different data for each layer. Here we use a filter (class = "subcompact") to select a subset of the data and plot only that subset. 
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```
Excersize: 
1. What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
* for a line char: geom_line()
* for a boxplot: geom_boxplot()
* for a histogram: geom_histogram()
* for an area chart: geom_ribbon()
  
2. Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_line()
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot()
ggplot(data = mpg, mapping = aes(x = hwy)) + 
  geom_histogram()
ggplot(data = mpg) + 
  geom_ribbon(aes(displ, ymin = 0, ymax = hwy, fill = drv, color = drv))

```
3. Will these two graphs look different? Why/why not?
* No they will look the same, because the second one is basically just hte duplicated version of the first one whic is globalized but they hold the exact same information. 

## Transformation and Stats
The diamonds dataset comes in ggplot2 and contains information about ~54,000 diamonds, including the price, carat, color, clarity, and cut of each diamond.

### Plotting Statistics 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```
More diamonds are available with high quality cuts than low quality cuts 

Geoms and stats can be used interchangebly
```{r}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```
### Overriding defaults in ggplot2
Change the default stat (which is a count, a summary) to identity (which is the raw value of a variable).
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```
You can also override a default mapping from transformed variables to aesthetics. For instance, you could display a bar chart of proportion of your total diamond dataset rather than a count.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
```
### Plotting Statistical Details 
Use stat_summary to show a lot more detail about transformations in your plot. 
```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```
### Positional Adjustments 
You can use colour or fill to change aspects of bar colours

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
#colour = cut only outlines the bar chart whereas fill = cut fills the entire bar
```
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```
If you use position = "identity", you will be able to place each object exactly where it falls in the context of the graph. This is vital for point charts like scatter plots but makes a mess in a bar plot situation by showing too much information (a bar plot generally summarises information). We will need to alter the bar aesthetic.
```{r}
#To alter transparency (alpha)
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")

#To color the bar outlines with no fill color
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```
Position = "fill" works like stacking, but makes each set of stacked bars the same height.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```
Position = "dodge" places overlapping objects directly beside one another.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```
Position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```
## The Layered Grammer of Graphics 

### Template 
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <FACET_FUNCTION>
  
## Assignment: Plot Deconstruction 


# Workshop 2 – Using ggplot2 for communication

###Labels 
In ggplot2, you add labels with the labs() function. 
Titles 
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")
```

Subtitle adds additional detail in a smaller font beneath the title and caption adds text at the bottom right of the plot 
Caption adds text at the bottom right of the plot, often used to describe the source of the data.
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```
You can also use labs() to replace axis labels and legend titles
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type"
  )
```
### Annotations
you can use geom_text() to add textual labels to your plots. This works in the same way as geom_point() but rather than a shape geometry it can add a label
```{r}
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class)
```
###Scales
```{r}
#ggplot automatically adds default scales 
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))
```
If you want to tweak the default scales you can do it by offering values to the scale parameters
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous("displ", 3) +
  scale_y_continuous("hwy", 2) +
  scale_colour_discrete("class", 9)
#no clue what is happening, it is just chaning the numbers 
```
###Axis Ticks 
You can also change the ticks on your axes. Breaks controls the position of the ticks and you can use labels as an argument to change the text label associated with ticks
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))
#this changes the spacing on the axis, first number is the start number, second is the max number and by is how much it is increasing by 

seq(15, 40, by = 5)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL)
# you can alos use labels = NULL to surpress the labels altogether
```
### Legends and Colour Schemes
Sometimes you might want to change the position of your legend, perhaps to make the plot itself as wide as possible (e.g. put the legend underneath) or to hide it all together. 

To control the overall position of the legend, you need to use a theme() setting. You can also use legend.position = "none" to suppress the display of the legend altogether
```{r}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right") # the default
#code for positioning the legend in different places. Create the base then add it to the theme 
```
###Replacing a Scale
There are two types of scales you’re mostly likely to want to switch out: continuous position scales and colour scales.
It’s very useful to plot transformations of your variable
```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10()
```
Another scale that is frequently customised is the colour scale. In fact, from a data visualisation point of you, this one is among the most important changes you can make to a base ggplot.
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1")
```
If there are just a few colours, you can add a redundant shape mapping. This will also help ensure your plot is interpretable in black and white.
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_colour_brewer(palette = "Set1")
```
When you have a predefined colours you want to use you can set them yourself, using scale_colour_manual()
```{r}
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue"))
```

To use one of the most popular colour schemes, simply use scale_colour_viridis() provided by the viridis package
```{r}
install.packages('viridis')
install.packages('hexbin')
library(viridis)
library(hexbin)

df <- tibble( # note we're just making a fake dataset so we can plot it
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + # a new geom!
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```
### Themes
Themes allow you to change some or all of the non-data elements of your plot with a theme.
ggplot2 has eight themes by default.
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light()

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic()

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark()
```
```{r}
theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Arial"))
```
###Saving and exporting your data 
In general, the best way to make a single file will be to export it using the ggsave() function. 
There are two main ways to get your plots out of R using ggplot2. 
ggsave() will save the most recent plot to your working directory
```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point()

ggsave("my-plot.pdf")
#> Saving 7 x 4.32 in image
```
# Workshop 3

```{r}
#install.packages("usethis")
#library(usethis)
use_git_config(user.name = "Jadeasiu", user.email = "jadeasiu@hotmail.com")
gitcreds::gitcreds_set()
```
Assignment 1
```{r}
install.packages("wesanderson")
library("wesanderson")

zooxanthellae <- read.csv("zooxanthellae.csv")
zooxanthellae

theme(panel.border = element_blank(),
      plot.title = element_text(hjust=0.5),
      plot.subtitle = element_text(hjust=0.5),
      legend.position="right",
      legend.text=element_text(size=8),
      legend.key = element_blank(),
      legend.background = element_blank(),
      axis.text.y=element_text(colour="black"),
      axis.text.x=element_text(colour="black"),
      text=element_text(family="Calibri"))

Age <- as.character(zooxanthellae$Age)
Age <- factor(Age, levels = c("4 weeks", "9 weeks", "18 weeks", "37 weeks", "Adults"))

ggplot(data = zooxanthellae) +
  geom_bar(
    mapping = aes(x = Age, y = Occurrence, fill = Strain), 
    stat = "identity",
    position = "stack"
  ) +
  labs(
    title = "Abundance of two Symbiodinium strains in A. Tenuis",
    subtitle = "Magnetic Island, QLD",
    color = "Sybiodinium Strain"
  ) +
  scale_y_continuous(breaks = seq(0, 100, by = 10), expand=c(0,0), limits=c(0,100)) +
  scale_fill_manual(values = wes_palette("Moonrise3")) +
  theme(plot.title = element_text(hjust=0.5),
        plot.subtitle = element_text(hjust=0.5),
        panel.background = element_blank(),
        legend.position="right",
        legend.text=element_text(size=8),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        axis.line = element_line(colour = "black"),
        text=element_text(family="Calibri"))
```

```{r}
ggplot(data = mpg) + 
+   geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

# Workshop 4: Data Wrangling in R

```{r}
vignette("tibble")
library(tidyverse)
iris
str(iris) #data fram with numerical data and 1 factor variable (species) with 3 levels
as_tibble(iris)
```
```{r}
tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y) # call new variables to produce new column values!
```
With tibble, you can include nonsyntactic variables, however you’ll need to refer to them with backticks to help R understand what’s going on. Note: You’ll need the backticks for these variables when using other packages as well (ggplot2, dplyr, tidyr etc.) so it can cause some issues but it can still be used as an option
```{r}
tb <- tibble(
  `:)` = "smile", 
  ` ` = "space",
  `2000` = "number"
)
tb
#> # A tibble: 1 × 3
#>   `:)`  ` `   `2000`
#>   <chr> <chr> <chr> 
#> 1 smile space number
```
### Tribbles: Transposed tibbles 
The purpose is tribbles is mainly only to help you do data entry directly in your script
  * The column headings are defined by formulas (they start with ~), and each data entry is put     in a column, separated by commas.
  
```{r}
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
#> # A tibble: 2 × 3
#>   x         y     z
#>   <chr> <dbl> <dbl>
#> 1 a         2   3.6
#> 2 b         1   8.5
```
### The difference between tibble and data.fram
The first difference lies in printing. Tibbles only print the first 10 rows and all columns that can fit on the screen making it easy to work with really large data.

```{r}
as_tibble(iris)
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)

```
Sometimes your may need to see more than the default 10 rows of data. You can use print() to designate the number of rows (n) and display width. (width = Inf #displays every column).

```{r}
install.packages("nycflights13")
library(nycflights13)
nycflights13::flights %>% 
  print(n = 10, width = Inf)
```
Some more ways you can interrogate your data with tibble: 
  * : if more than n rows, print only m rows. 
  * Use options(tibble.print_min = Inf) to always show all rows.
  * Use options(tibble.width = Inf) to always print all columns, regardless of the width of the     screen.
  
```{r}
options(tibble.width = Inf) #from now on you will always be able to see all of your coloumns when using tibble 
```

### Pulling out a single variable from our dataframe
We can use a dollar sign ($) to extract a full column of data (by name), or the simple double bracket to pull out an exact row of data (by row position). You can also pull out an exact cell value using [[,]]

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)
#Extract name 
df$x
df[["x"]]

#Extract by row position 
df[[1]]

# Extract by exact position
df[[2,2]]
```

### Using Pipes
To use these base R functions we will need to use a  . as a placeholder when using pipes to use these functions

```{r}
df %>% .$x
df %>% .[["x"]]
```
Tibble will not do partial matchings, which means if the variable you call out doesn’t match exactly what’s in the dataframe, and you will get a warning sign 
```{r}
df <- tibble(
    xxx = runif(5),
    y = rnorm(5)
)
df$xx
NULL
```
```{r}
df <- data.frame(abc = 1, xyz = "a")
df
df$x #call by name
df[, "xyz"] #call by exact position

#In this example, we made a dataframe with 2 coloumns, 'abc' and 'xyz', each with 1 row. When calling x by name, R thinks that you want what is in coloumn 'xyz' and therefore gives you 'a', which it also does the same then when calling its exact position. 
```

## How can I import data?

  *read_csv() reads comma delimited files, read_csv2() reads semicolon separated files (common    in countries where , is used as the decimal place), read_tsv() reads tab delimited files, and   read_delim() reads in files with any delimiter.
  * read_fwf() reads fixed width files. You can specify fields either by their widths with        fwf_widths() or their position with fwf_positions(). read_table() reads a common variation of   fixed width files where columns are separated by white space.
  
```{r}
read_csv("a,b,c
1,2,3,
4,5,6")
```
read_csv() uses the first line of data for column names.  if you have metadata at the top of your file, you might want to skip these lines using skip = n where n is the number of lines you want to skip over. 
Alternatively, use comment = '#' to will drop all lines starting with a “#” or whatever character you designate.
```{r}
read_csv("The first line of metadata
  The second line of metadata
  x,y,z
  1,2,3", skip = 2)
```
```{r}
read_csv("# A comment I want to skip
  x,y,z
  1,2,3", comment = "#")
```
Another reason to adjust the read_csv() default is if your data does not contain column names. In this case, use col_names = FALSE to tell read_csv() not to treat the first row as headings but to instead label them sequentially from X1 to Xn. 
```{r}
read_csv("1,2,3\n4,5,6", col_names = FALSE)
#Here ("\n" is simply a shortcut for adding a new line. This is a common ‘break’ in programming.
```
You can pass col_names a character vector to be used as the column names:
```{r}
read_csv("1,2,3\n4,5,6", col_names = c("x", "y", "z"))
```
You can also set the no data values. This is important because if you assume an empty cell is a 0, you could make a grave error in your analysis. Better to tell R that this is ‘no data’ so you can easily remove it later on. Sometimes people even use weird numbers for no data values, like 9999. Here you can tell R that this is not a real number, rather code for ‘no data’.
Another option that commonly needs tweaking is na: this specifies the value (or values) that are used to represent missing values in your file:
```{r}
read_csv("a,b,c\n1,2,.", na = ".")
```

## Tidying data using Tidyr
How we make our dataset a tidy dataset is by following three interrelated rules. 
  *Each variable must have its own column.
  *Each observation must have its own row.
  *Each value must have its own cell.

```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)
```
```{r}
# Compute cases per year
table1 %>% 
  count(year, wt = cases)
```
```{r}
# Visualise changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```
Common problems with untidy data:
  *One variable is spread across multiple columns
  *One observation is scattered across multiple rows
We can use pivot_longer() and pivot_wider() to help fix these problems 

A common problem is a dataset where some of the column names are not names of variables, but values of a variable. 
To tidy a dataset like this, we need to pivot the offending columns into a new pair of variables. To describe that operation we need three parameters:
  *The set of columns whose names are values, not variables. In this example, those are the       columns 1999 and 2000.
  *The name of the variable to move the column names to. Here it is year.
  *The name of the variable to move the column values to. Here it is cases.
Together those parameters generate the call to pivot_longer():

```{r}
table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
```
pivot_longer() makes datasets “longer” by increasing the number of rows and decreasing the number of columns.
We can use pivot_longer() to tidy table4b in a similar fashion. The only difference is the variable stored in the cell values:
```{r}
table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
```
To combine the tidied versions of table4a and table4b into a single tibble, we need to use dplyr::left_join(), which we learn about in the next section.

```{r}
tidy4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
left_join(tidy4a, tidy4b)
```
Using pivot_wider():
To handle an observation scattered across multiple rows. pivot_wider() is the opposite of pivot_longer(). In table2 we see an observation is a country in a year with the observation spread across two rows:
```{r}
table2
```
To make this tidy we only need two parameters:
  *The column to take variables from: type
  *The column to take values from: count
Now we can use pivot_wider() to make the table shorter and wider by creating new, separate columns for cases and population and populating them with their associated values.

```{r}
table2 %>%
    pivot_wider(names_from = type, values_from = count)
```
### Separating and uniting data tables

In table3, we see one column (rate) that contains two variables (cases and population). To address this, we can use the separate() function which separates one column into multiple columns wherever you designate.
```{r}
table3
```
As we said above, we need to split the rate column up into two variables. separate() will take the name of the column to split and the names of the columns we want it split into.
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))
```
By default, separate() will split values wherever it sees a non-alphanumeric character (i.e. a character that isn’t a number or letter). For example, in the code above, separate() split the values of rate at the forward slash characters. If you wish to use a specific character to separate a column, you can pass the character to the sep argument of separate(). For example, we could rewrite the code above as:
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")
```
Notice the data types in table3 above. Both cases and population are listed as character types. This is a default of using separate(). However, since the values in those columns are actually numbers, we want to ask separate() to convert them to better types using convert = TRUE
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
```
We can also perform the inverse of separate() by using unite() to combine multiple columns into a single column. In the example below for table5 we use unite() to rejoin century and year columns. unite() takes a data frame, the name of the new variable and a set of columns to combine using dplyr::select(). 
```{r}
table5 %>% 
  unite(new, century, year, sep = "")
```
### Handling missing values 

In the below example, the return for the 4th quarter of 2015 is missing as NA while the return for the 1st quarter of 2016 is missing because it is simply absent.
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
stocks
```
```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return)
```
Because these explicit missing values may not be important in other representations of the data, you can set values_drop_na = TRUE in pivot_longer() to turn explicit missing values implicit. 
This makes those missing values that are probably not supposed to be missing now a valid row of data in your data frame
```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(
    cols = c(`2015`, `2016`), 
    names_to = "year", 
    values_to = "return", 
    values_drop_na = TRUE
  )
```

Another important tool for making missing values explicit - very clear to you that they represent actual missing data values -  in tidy data is complete(). This function takes a set of columns and finds all the unique combinations and then ensures the original dataset contains all of those values, including filling in explicit NA where needed.
```{r}
stocks
stocks %>% 
  complete(year, qtr)
```
The fill() function can be used to fill in missing values that were meant to be carried forward in the data entry process. It can take columns with missing values and carry the last observation forward (replace them with the most recent non-missing value).
```{r}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
treatment

treatment %>% 
  fill(person)
```
# Workshop 5: Spatial Data in R

```{r}
#copepods_raw.csv
#install.packages("sf") 
#install.packages("terra")
#install.packages("tmap")

#load into R library
library(tidyverse)
library(sf) # simple features
library (terra) # for raster
library(tmap) # Thematic maps are geographical maps in which spatial data distributions are visualized
```
```{r}
#load the copepod data into R studio
library(readr)
dat <- read_csv("data-for-course/copepods_raw.csv")
dat
```
Plot the coordinates for the samples (segments of the CPR silks)
```{r}
library(ggplot2)
ggplot(dat) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()
```
Look at richness in a non-spatial way
```{r}
ggplot(dat, aes(x = latitude, y = richness_raw)) + 
  stat_smooth() + 
  geom_point()
```
Turn our point data into a spatially referenced data frame using the sf package (sf stands for ‘simple features’) which is an open standard for geospatial databases
  *st_as_sf converts different data types to simple features. 
  *dat is our original data. 
  *coords gives the names of the columns that relate to the spatial coordinates (in order of X    coordinate followed by Y coordinate).
  *crs stands for coordinate reference system which we will discuss next.
```{r}
sdat <- st_as_sf(dat, coords = c("longitude", "latitude"), 
                 crs = 4326)
```

### Coordinate Reference Systems
Coordinate reference systems are required for 2D mapping to compensate for the lumpy, spherical (3D) nature of the earth
In R, best practice is to either use an EPSG code or Well-known text (WKT) to define a CRS. A WKT string contains all of the detailed information we need to define a crs, but is cumbersome if you don’t need all of the detail

```{r}
crs4326 <- st_crs(4326)
crs4326 # look at the whole CRS
crs4326$Name # pull out just the name of the crs
crs4326$wkt # crs in well-known text format
```
### Cartography 
```{r}
plot(sdat["richness_raw"])
```
Here we have only plotted the richness column. If we used plot(sdat) it would create a panel for every variable in our dataframe. In sf, we can use square brackets ["richness_raw"] to select a single variable.

```{r}
plot(sdat)
```

### Thematic Maps for Communication 
tmap works similarly to ggplot2 in that we build and add on layers. Here we only have one layer from sdat. We declare the layer with tm_shape() (in this case sdat), then the plot type with the following command
  *tm_dots to plot dots of the coordinates. Other options are tm_polygons, tm_symbols and many    others we’ll see later.
  *We’ve chosen "richness_raw" as the color scale


```{r}
#using tmap

tm1<-tm_shape(sdat) + 
  tm_dots(col = "richness_raw")
tmap_save(tm1, filename = "Richness-map.png", 
          width = 600, height = 600)
```

###Mapping spatial polygons as layers
The polygons in this copepod data are stored as shapefiles. 
Note that .shp files are generally considered an undesirable file format because they are inefficient at storing data and to save one shapefile you actually create multiple files. This means bits of the file might be lost if you transfer the data somewhere else
A better format than shapefile is the Geopackage which can save and compress multiple different data types all in a single file
We can read shapefiles directly into R with the st_read command (which is like read_csv, but for spatial files):
```{r}
aus <- st_read("data-for-course/spatial-data/Aussie/Aussie.shp")
shelf <- st_read("data-for-course/spatial-data/aus_shelf/aus_shelf.shp")
aus
```
```{r}
tm_shape(shelf) + 
  tm_polygons()
```
Here we have indicated the shape of our map (shelf) and we have added a command bbox = sdat to expand the extent of the map so it depicts all of our copepod data points. We then add the shape of Australia (aus) on top of the shelf, and finally our copepod data (sdat) in the form of points using tm_dots().
```{r}
tm_shape(shelf, bbox = sdat) + 
  tm_polygons() +
  tm_shape(aus) + 
  tm_polygons() + 
  tm_shape(sdat) + 
  tm_dots()+
  tmap_style("beaver")
```
### Exploring t_map
```{r}
vignette('tmap-getstarted')
library(tmap)
data("World")
tmap_mode("view")
tm_shape(World) +
    tm_polygons(c("HPI", "economy")) +
    tm_facets(sync = TRUE, ncol = 2)
```
```{r}
tmap_mode("plot")
## tmap mode set to plotting

data(NLD_muni)

NLD_muni$perc_men <- NLD_muni$pop_men / NLD_muni$population * 100

tm_shape(NLD_muni) +
    tm_polygons("perc_men", palette = "RdYlBu") +
    tm_facets(by = "province")
```
```{r}
tmap_mode("plot")
## tmap mode set to plotting

data(NLD_muni)
tm1 <- tm_shape(NLD_muni) + tm_polygons("population", convert2density = TRUE)
tm2 <- tm_shape(NLD_muni) + tm_bubbles(size = "population")

tmap_arrange(tm1, tm2)
```
```{r}
tmap_mode("view")
tm_basemap("Stamen.Watercolor") +
tm_shape(metro) + tm_bubbles(size = "pop2020", col = "red") +
tm_tiles("Stamen.TonerLabels")
```
```{r}
tmap_options(bg.color = "black", legend.text.color = "white")

tm_shape(World) +
    tm_polygons("HPI", legend.title = "Happy Planet Index")
```
###Exporting your map
using tmap_save() to save your map as either an .html file or an image file (.JPG, .PNG etc.) to your designated folder.   



